Mixture Models for Single-Cell Assays with Application to Vaccine Studies

### 2012-07-29 MIMOSA.tex
Minor changes.

\title[MIMOSA: Combinatorial Mixture Models or Single-Cell Assays]{Combinatorial Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
%\author[1]{Greg Finak}
%\author[2]{SC De Rosa}
%\author[3]{Mario Roederer}
%\author[1]{Raphael Gottardo}

@@

\begin{abstract}
Most biological samples (\textit{e.g.}, blood and tissue) are composed of many functionally distinct cell subsets that can only be measured accurately using single-cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on single-cell assays to provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combination) in the cell subset(s) of interest.
Here we present a Bayesian hierarchical framework based on beta-binomial mixture models for testing for differential marker expression using such single-cell assays. Our model allows the inference specific to each observation, while borrowing strength across observations through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental immunological assays measuring proteins or genes at the single-cell level, we find that our method has higher sensitivity and specificity than alternative methods. Using a simulation study we show that our framework is also robust to model misspecification. Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a multinomial-Dirichlet model. 
\end{abstract}

\begin{keywords}
Mixture modelling, hierarchical modelling, bayesian modelling, single-cell assays, immunology
\end{keywords}

@@

\section{Introduction}
\label{s:intro}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them. 
This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures. 
For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins. 
In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). 
Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular markers (up to 17 markers). 
More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}. 
Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells. 
A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}. 
While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}. 
This heterogeneity may carry important information, thus single-cell expression data should not be analyzed in the same fashion as cell-population level data. 
Special treatment of single-cell level data, which preserves information about population heterogeneity, is warranted in general. 
For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. 
These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood cell mononuclease (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

@@

The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine. 
Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8), used to subset T-cells, and functional markers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected. 
Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted. 
These counts are compared between antigen stimulated and unstimulated samples from an individual to identify significant differences. individuals who generate a response after stimulation are called \textit{responders} whereas individual that do not show any differences are called \textit{non-responders}. 
In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.

Although there is no standard approach to analyzing single-cell assays current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej,Proschan:2009ks}. 
All of these methods test observations separately, and no information is shared across observations even though one could expect some similarities across responders (\textit{resp.} non-responders). %In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}\footnote{What are we trying to say here? Let's update this sentence later}.

The framework developed in this paper addresses these issues explicitly. In our model, cell counts are modelled by a binomial (or multinomial in the multivariate case) distribution and information is shared across individuals by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood. 
In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across individuals. 
This sharing of information across individuals helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increase sensitivity and specificity when detecting responders. 
Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}.

>>>>

\title[MIMOSA: Mixture Models or Single-Cell Assays]{Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
%\author[1]{Greg Finak}
%\author[2]{SC De Rosa}
%\author[3]{Mario Roederer}
%\author[1]{Raphael Gottardo}

@@

\begin{abstract}
Most biological samples (\textit{e.g.}, blood and tissue) are composed of many functionally distinct cell subsets that can only be measured accurately using single-cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on single-cell assays to provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combination) in the cell subset(s) of interest.
Here, we present a Bayesian hierarchical framework based on a beta-binomial mixture model for testing for differential marker expression using single-cell assays. Our model allows the inference to be individual specific, while borrowing strength across individuals through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental assays measuring proteins or genes at the single-cell level and simulated data, we show that our method has higher sensitivity and specificity than alternative methods. Additional simulations show that our framework is also robust to model misspecification. Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a Dirichlet-multinomial model and illustrate this multivariate approach using single-cell gene expression data and simulations.
\end{abstract}

\begin{keywords}
Bayesian Inference, Expectation-Maximization, Flow cytometry, Gene expression, Hierarchical modeling, Immunology, Markov chain Monte Carlo
\end{keywords}

@@

\section{Introduction}
\label{s:intro}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them. 
This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures. 
For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins. 
In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). 
Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular markers (up to 17 markers)\footnote{Reference Mario's paper}. 
More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}. 
Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells. 
A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}. 
While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}. 
This heterogeneity may carry important information, thus single-cell expression data should not be analyzed in the same fashion as cell-population level data. 
Special treatment of single-cell level data, which preserves information about population heterogeneity, is warranted in general. 
For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. 
These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood, peripheral blood cell mononuclease (PBMC), or pre-sorted cell populations (\textit{e.g.} T-cells), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

@@

The ICS assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring cytokine production in response to activation following stimulation by an antigen that closely matches what was present in the original vaccine. 
Individual cells are labelled using fluorescently conjugated antibodies against phenotypic markers (CD3, CD4, and CD8), used to subset T-cells, and functional markers (cytokines) used to define antigen specific T-cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
A sufficiently large number of cells must be collected to ensure that the rare cell populations can be detected. 
Subsequently, each individual cell is classified as either positive or negative for each maker based on predetermined thresholds, then the number of cells matching each subpopulation phenotype is counted. 
These counts are compared between antigen stimulated and unstimulated samples from an individual to identify significant differences. Individuals who generate a response after stimulation are called \textit{responders} whereas individual that do not show any differences are called \textit{non-responders}. 
In many immunological studies, the size of the functionally distinct subpopulations (\textit{i.e.}, the number of positive cells) is very low (relative to the total number of cells), and real biological differences might be difficult to detect.

Although there is no standard approach to analyzing single-cell assays current methods range from ad-hoc rules based on log-fold changes, to permutation tests based on Hotelling's T$^2$ statistics, to exact tests of 2x2 contingency tables (\textit{e.g.}, Fisher's exact test and $\chi^2$ test)~\citep{Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Nason:2006dx,Peiperl:2010ej,Proschan:2009ks}. 
All of these methods test individuals separately, and no information is shared across individuals\footnote{We should say at some point that here we use individuals but the approach is general, could be used with animals, for example} even though one could expect some similarities across responders (\textit{resp.} non-responders). %In addition, these methods generally test one marker at a time or pairwise combinations of markers, raising questions of appropriate multiple testing adjustments, or they perform global tests of significance on multiple markers resulting in decreased power to detect small changes in subsets of cytokines~\cite{Proschan:2009ks,Nason:2006dx}\footnote{What are we trying to say here? Let's update this sentence later}.

The framework developed in this paper addresses these issues explicitly. In our model, cell counts are modelled by a binomial (or multinomial in the multivariate case) distribution and information is shared across individuals by means of a prior distribution placed on the unknown proportion(s) of the binomial (or multinomial) likelihood. 
In order to discriminate between responders and non-responders, the prior is written as a mixture of two beta (or Dirichlet in the multivariate case) distributions where the hyper-parameters for each mixture component are shared across individuals. 
This sharing of information across individuals helps regularize proportion estimates when the cell counts are small, which is typical with single-cell assays, and increase sensitivity and specificity when detecting responders. 
Because our framework is multivariate in nature, multiple cell subsets can be model simultaneously, which could help detect small biological changes that are spread out across multiple cell subsets~\citep{Nason:2006dx}. Our paper is organized as follows, ...
